\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-11 k-planar.tex
Always compile before committing!
Pfffft

@@

% \referee{2}{Page 3. $P$ is used twice in the statement of Theorem 3.  It is probably
safer (and necessary?) to use a different variable for the second
occurrence.}

@@

\note{DW}{Use $\SS$ for the given shortcut system and $\PP$ for the partition?}
\note{PM}{Yes, I've done this.  Again, I was carefuly, but need to reread.}

@@

We first formally define $k$-planar graphs.  An \defin{embedded graph} $G$ is a graph with $V(G)\subset\R^2$ in which each edge $vw\in E(G)$ is a curve\footnote{A \defin{curve} in a surface $\Sigma$ is a continuous function $f:[0,1]\to \Sigma$. The points $f(0)$ and $f(1)$ are called the \defin{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as the point set $\{f(t):0\le t\le 1\}$.} in $\R^2$ with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior.  A \defin{crossing} in an embedded graph $G$ is a triple $(p,vw,xy)$ with $p\in\R^2$, $vw,xy\in E(G)$ and such that $p\in (vw\cap xy)\setminus\{v,w,x,y\}$. An embedded graph $G$ is \defin{$k$-plane} if each edge of $G$ takes part in at most $k$ crossings. \note{DW}{We need to assume that at most two edges cross at a single point, so we can add a dummy vertex of degree 4, right?}\note{PM}{No, under the current definition, $r$ edges crossing at a single point $p$ generate $\binom{r}{2}$ crossings, with each edge taking part in $r-1$ crossings, just like they would if we perturbed the edges so that each pair crosses at a distinct point.  I guess we could mention this.  (See the first paragraph in the proof of Theorem~2 below.)} \note{DW}{The downside of this approach is that in the opening section we give an informal definition of $k$-planar, which needs the assumption of having no three edges cross at a point, since we use this assumption in the construction of a $(k+1,2)$-shortcut system for $k$-planar graphs give in the Introduction. I think we should not have different definitions in two parts of the paper.} A (not necessarily embedded) graph $G'$ is \defin{$k$-planar} if there exists a $k$-plane graph $G$ isomorphic to $G'$. Under these definitions, $0$-planar graphs are exactly planar graphs and $0$-plane graphs are exactly plane graphs.

@@

\begin{proof}[Proof of \cref{kPlanarProduct}]
	Let $G$ be a $k$-plane graph.  We assume, for ease of exposition, that any point $p\in\R^2$ is involved in at most one crossing $(p,vw,xy)$ of $G$. This assumption is justified since it can be enforced by a slight deformation of the edges of $G$ and the resulting (deformed) graph is also $k$-plane.

	As in the proof of \cref{AddDummy}, let $G_0$ be the plane graph obtained by adding a dummy vertex at each crossing in $G$. In this way, each edge $vw\in E(G)$ corresponds naturally to a path $P_{vw}$ of length at most $k+1$ in $G_0$.  Let $\SS := \{P_{vw}: vw\in E(G)\}$. Observe that $\SS$ is a $(k+1,2)$-shortcut system for $G_0$ and that $G_0^{\SS}\supseteq G$.  Specifically, $G_0^{\SS}$ contains every edge and vertex of $G$ as well as the dummy vertices in $V(G_0)\setminus V(G)$ and their incident edges.

@@

\note{PM}{Include in note to referee:  We choose to keep this section because
  $1$-planar graphs are of special interest to the graph drawing community.  For example, the annotated bibliography on $1$-planarity (\htmladdnormallink{https://arxiv.org/abs/1703.02261}{arXiv:1703.02261}) has 143 entries, and even this is now four years out of date.

  An important aspect of our result for $1$-planar graphs is that graph $H$ in the product is planar and has treewidth $3$, i.e., it has simple treewidth $3$.  In a more recent application of product structure \citet{BDJM} show that such graphs have $\ell$-vertex rankings using $O(\log n/\log\log\log n)$ colours, and this is tight. This result requires that $H$ have simple treewidth $3$, even treewidth $3$ is not sufficient.
}

\note{DW}{I will rewrite the intro to this subsection once we decide what our results are.}

@@

Let $G$ be a $1$-plane graph.  We make $G$ into an \defin{edge-maximal} $1$-plane multigraph, by repeating the following operation:  If two vertices $v$ and $w$ of $G$ appear on a common face $F$ and there is no edge $vw\in E(G)$ contained in the boundary of $F$, then add $vw$, embedded in the face $F$.  This may introduce parallel edges into $G$, but no two such edges will appear on the boundary of a common face.

@@

To understand the structure of $G$, it is helpful to consider the plane graph $G_0$ obtained by replacing each pair of edges $vw$ and $xy$ involved in a crossing $(p,vw,xy)$ with four edges $vp$, $wp$, $xp$, and $yp$ meeting at a newly added \defin{dummy vertex} $p$. Let $F$ be a face of $G_0$ and let $W:=w_0,\ldots,w_{r-1}$ be the facial walk around $F$. In the following paragraphs, subscripts on the vertices in $W$ are implicitly taken modulo $r$. By edge maximality, if $W$ contains only non-dummy vertices, then it contains exactly three vertices and $F$ is bounded by three edges of $G$.

@@

Consider some crossing $(p,vw,xy)$ in $G$.  The point $p$ at which this crossing occurs is a vertex of $G_0$ that has degree $4$.  There are four faces $F_1,\ldots,F_4$ of $G_0$ each with $p$ on their boundary.  Each of these four faces of is of Type~(ii), so the boundary of $F_1\cup\cdots\cup F_4$ includes a $4$-cycle $C:=vxwy$ in $G$ whose edges are not involved in any crossings.

>>>>

\subsection{$k$-Planar Graphs}
\label{k_planar_section}

@@

% \referee{2}{Page 3. $P$ is used twice in the statement of Theorem 3.  It is probably
% safer (and necessary?) to use a different variable for the second
% occurrence.}

@@

% \note{DW}{Use $\SS$ for the given shortcut system and $\PP$ for the partition?}
% \note{PM}{Yes, I've done this.  Again, I was carefuly, but need to reread.}

@@

We first formally define $k$-planar graphs.  An \defin{embedded graph} $G$ is a graph with $V(G)\subset\R^2$ in which each edge $vw\in E(G)$ is a curve\footnote{A \defin{curve} in a surface $\Sigma$ is a continuous function $f:[0,1]\to \Sigma$. The points $f(0)$ and $f(1)$ are called the \defin{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as the point set $\{f(t):0\le t\le 1\}$.} in $\R^2$ with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior.  A \defin{crossing} in an embedded (multi)graph $G$ is a pair of distinct edges $vw,xy\in E(G)$ whose interior have a point in common. Without any loss of generality we may assume that the interiors of any two edges of $G$ have at most one point in common and that no point in $\R^2$ is contained in the interior of three or more edges of $G$.  This assumption can be enforced by a local change that does not change the number of crossings each edge is involved in.  If $p$ is a common point in the interior of two crossing edges $vw$ and $xy$ then we say that $vw$ and $xy$ \defin{cross at} $p$.

% (This assumption is already present in the informal definition of $k$-planar graphs used in \cref{k_planar_section}.)

%
% \note{DW}{We need to assume that at most two edges cross at a single point, so we can add a dummy vertex of degree 4, right?}\note{PM}{No, under the current definition, $r$ edges crossing at a single point $p$ generate $\binom{r}{2}$ crossings, with each edge taking part in $r-1$ crossings, just like they would if we perturbed the edges so that each pair crosses at a distinct point.  I guess we could mention this.  (See the first paragraph in the proof of Theorem~2 below.)} \note{DW}{The downside of this approach is that in the opening section we give an informal definition of $k$-planar, which needs the assumption of having no three edges cross at a point, since we use this assumption in the construction of a $(k+1,2)$-shortcut system for $k$-planar graphs give in the Introduction. I think we should not have different definitions in two parts of the paper.} A (not necessarily embedded) graph $G'$ is \defin{$k$-planar} if there exists a $k$-plane graph $G$ isomorphic to $G'$. Under these definitions, $0$-planar graphs are exactly planar graphs and $0$-plane graphs are exactly plane graphs.

@@

\begin{proof}[Proof of \cref{kPlanarProduct}]
	Let $G$ be a $k$-plane graph.
    % We assume, for ease of exposition, that any point $p\in\R^2$ is involved in at most one crossing $(p,vw,xy)$ of $G$. This assumption is justified since it can be enforced by a slight deformation of the edges of $G$ and the resulting (deformed) graph is also $k$-plane.
	As in the proof of \cref{AddDummy}, let $G_0$ be the plane graph obtained by adding a dummy vertex at each crossing in $G$. In this way, each edge $vw\in E(G)$ corresponds naturally to a path $P_{vw}$ of length at most $k+1$ in $G_0$.  Let $\SS := \{P_{vw}: vw\in E(G)\}$. Observe that $\SS$ is a $(k+1,2)$-shortcut system for $G_0$ and that $G_0^{\SS}\supseteq G$.  Specifically, $G_0^{\SS}$ contains every edge and vertex of $G$ as well as the dummy vertices in $V(G_0)\setminus V(G)$ and their incident edges.

@@

% \note{PM}{Include in note to referee:  We choose to keep this section because
%   $1$-planar graphs are of special interest to the graph drawing community.  For example, the annotated bibliography on $1$-planarity (\htmladdnormallink{https://arxiv.org/abs/1703.02261}{arXiv:1703.02261}) has 143 entries, and even this is now four years out of date.
%
%   An important aspect of our result for $1$-planar graphs is that graph $H$ in the product is planar and has treewidth $3$, i.e., it has simple treewidth $3$.  In a more recent application of product structure \citet{BDJM} show that such graphs have $\ell$-vertex rankings using $O(\log n/\log\log\log n)$ colours, and this is tight. This result requires that $H$ have simple treewidth $3$, even treewidth $3$ is not sufficient.
% }
%
% \note{DW}{I will rewrite the intro to this subsection once we decide what our results are.}

@@

Let $G$ be a $1$-plane graph.  We make $G$ into an \defin{edge-maximal} $1$-plane multigraph, by repeating the following operation:  If two vertices $v$ and $w$ of $G$ appear on a common face $F$ and there is no edge $vw\in E(G)$ contained in the boundary of $F$, then add the edge $vw$, embedded in the face $F$.  This may introduce parallel edges into $G$, but no two such edges will appear on the boundary of a common face.

@@

To understand the structure of $G$, it is helpful to consider the plane graph $G_0$ obtained by replacing each pair of edges $vw$ and $xy$ that cross at $p$ with four edges $vp$, $wp$, $xp$, and $yp$ meeting at a newly added \defin{dummy vertex} $p$. Let $F$ be a face of $G_0$ and let $W:=w_0,\ldots,w_{r-1}$ be the facial walk around $F$. In the following paragraphs, subscripts on the vertices in $W$ are implicitly taken modulo $r$. By edge maximality, if $W$ contains only non-dummy vertices, then it contains exactly three vertices and $F$ is bounded by three edges of $G$.

@@

Consider some pair of edges $vw$ and $xy$ that cross at $p$.  The point $p$ is a vertex of $G_0$ that has degree $4$.  There are four faces $F_1,\ldots,F_4$ of $G$ each with $p$ on their boundary.  Each of these four faces of $G$ is of Type~(ii), so the boundary of $F_1\cup\cdots\cup F_4$ includes a $4$-cycle $C:=vxwy$ in $G$ whose edges are not involved in any crossings.
